Mechanized Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-01-18 paper-new/ordering.tex
changes to ordering.tex

As a first attempt at giving a semantics to the ordering, we could try to model types as
sets equipped with an ordering that models term precision. Since term precision is reflexive
and transitive, and since we identify terms that are equi-precise, we choose to model types
as partially-ordered sets. We model the term precision ordering $M \ltdyn N : A \ltdyn B$ as an
ordering relation between the posets denoted by $A$ and $B$.

However, it turns out that modeling term precision by a relation defined by guarded fixpoint
is not as straightforward as one might hope.
A first attempt might be to define an ordering $\semltbad$ between $\li X$ and $\li Y$
that allows for computations that may take different numbers of steps to be related.
The relation is parameterized by a relation $\le$ between $X$ and $Y$, and is defined
by guarded fixpoint as follows:
% simultaneously captures the notions of error approximation and equivalence up to stepping behavior:

@@

The problem with this definition is that the resulting relation is \emph{provably} not
transitive: it can be shown (in Clocked Cubical Type Theory) that if $R$ is a
relation on $\li X$ satisfying three specific properties, one of which is
transitivity, then that relation is trivial.
(The other two properties are that the relation is a congruence with respect to $\theta$,
and that the relation is closed under delays $\delta = \theta \circ \nxt$ on either side.)
Since the above relation \emph{does} satisfy the other two properties, we conclude
that it must not be transitive.

@@

We are therefore led to wonder whether we can formulate a version of the relation
that \emph{is} transitive.
It turns out that we can, by sacrificing another of the three properties from
the above lemma. Namely, we give up on closure under delays. Doing so, we end up
with a \emph{lock-step} error ordering, where, roughly speaking, in order for
computations to be related, they must have the same stepping behavior.
%
We then formulate a separate relation, \emph{weak bisimilarity}, that relates computations
that are extensionally equal and may only differ in their stepping behavior.

@@

  \item For a predomain $A$, there is a predomain $\li A$ for the ``lift'' of $A$
  using the lift monad. We use the same notation for $\li A$ when $A$ is a type
  and $A$ is a predomain, since the context should make clear which one we are referring to.
  The underling type of $\li A$ is simply $\li \ty{A}$, i.e., the lift of the underlying
  type of $A$.
  The ordering on $\li A$ is the ``lock-step error-ordering'' which we describe in
  \ref{subsec:lock-step}. The bismilarity relation is the ``weak bisimilarity''
  described in Section \ref{}

@@

\subsubsection{Step-Sensitive Error Ordering}\label{subsec:lock-step}

As mentioned, the ordering on the lift of a predomain $A$ is called the
\emph{step-sensitive error-ordering} (also called ``lock-step error ordering''),
the idea being that two computations $l$ and $l'$ are related if they are in
lock-step with regard to their intensional behavior, up to $l$ erroring.
Formally, we define this ordering as follows:

@@

We also define a heterogeneous version of this ordering between the lifts of two
different predomains $A$ and $B$, parameterized by a relation $R$ between $A$ and $B$.

\subsubsection{Weak Bisimilarity Relation}

@@

>>>>

As a first attempt at giving a semantics to the ordering, we could try to model types as
sets equipped with an ordering that models term precision. Since term precision is reflexive
and transitive, and since we identify terms that are equi-precise, we could model types
as partially-ordered sets and terms $\Gamma \vdash M : B$ as monotone functions.
We could then model the term precision ordering $M \ltdyn N : A \ltdyn B$ as an
ordering relation between the monotone functions denoted by $M$ and $N$.

However, it turns out that modeling term precision is not as straightforward as one might hope.
A first attempt might be to define an ordering $\semltbad$ between $\li X$ and $\li Y$
that allows for computations that may take different numbers of steps to be related.
The relation is parameterized by a relation $\le$ between $X$ and $Y$, and is defined
by guarded fixpoint as follows:
% simultaneously captures the notions of error approximation and equivalence up to stepping behavior:

@@

The problem with this definition is that the resulting relation is \emph{provably} not
transitive: it can be shown (in Clocked Cubical Type Theory) that if $R$ is a
relation on $\li X$ satisfying three specific properties, one of which is
transitivity, then that relation is trivial.
The other two properties are that the relation is a congruence with respect to $\theta$,
and that the relation is closed under delays $\delta = \theta \circ \nxt$ on either side.
Since the above relation \emph{does} satisfy the other two properties, we conclude
that it must not be transitive.

@@

We are therefore led to wonder whether we can formulate a version of the relation
that \emph{is} transitive.
It turns out that we can, by sacrificing another of the three properties from
the above result. Namely, we give up on closure under delays. Doing so, we end up
with a \emph{lock-step} error ordering, where, roughly speaking, in order for
computations to be related, they must have the same stepping behavior.
%
We then formulate a separate relation, \emph{weak bisimilarity}, that relates computations
that are extensionally equal and may only differ in their stepping behavior.

Finally, the semantics of term precision will be a combination of these two relations.
% TODO Define it here?

@@

  \item For a predomain $A$, there is a predomain $\li A$ for the ``lift'' of $A$
  using the lift monad. We use the same notation for $\li A$ when $A$ is a type
  and $A$ is a predomain, since the context should make clear which one we are referring to.
  The underling type of $\li A$ is simply $\li \ty{A}$, i.e., the lift of the underlying
  type of $A$.
  The ordering on $\li A$ is the ``lock-step error-ordering'' which we describe in
  \ref{sec:lock-step}. The bismilarity relation is the ``weak bisimilarity''
  described in Section \ref{}

@@

\subsubsection{Lock-Step Error Ordering}\label{sec:lock-step}

As mentioned, the ordering on the lift of a predomain $A$ is called the
\emph{lock-step error-ordering} (also called the ``step-sensitive error ordering''),
the idea being that two computations $l$ and $l'$ are related if they are in
lock-step with regard to their intensional behavior, up to $l$ erroring.
Formally, we define this ordering as follows:

@@

We similarly define a heterogeneous version of this ordering between the lifts of two
different predomains $A$ and $B$, parameterized by a relation $R$ between $A$ and $B$.

\subsubsection{Weak Bisimilarity Relation}\label{sec:weak-bisimilarity}

@@

% To manage this construction, we break it down into multiple steps and do it modularly

% We're not proving that the term model satisfies graduality 
% (ex. the exponential in the term model includes all functions, whereas the
% predomain model only includes monotone functions)
